---
title: "Worksheet 10: Advanced Data Manipulation"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE, 
                      warning = TRUE,
                      message = FALSE,
                      fig.align = "center", 
                      R.options = list(max.print=50))

# Edit the file starting below

```

In this worksheet, you will practice more advanced examples of using `dplyr` functions to manipulate our data.

## 1. Set up and introduction to the dataset

Let's load `tidyverse` which contains the `dplyr` package:

```{r}
# Load a package
library(tidyverse)
```

We will continue using the dataset `txhousing` for this worksheet:

```{r}
# Take a look at the first few rows of the dataset
head(txhousing) 
```

## 2. More examples of creating new variables

We can create new variables based on some conditions on the values of some variables. For example, recall `ifelse()`:

```{r}
# Create a new variable under a condition with ifelse()
txhousing |>
  mutate(semester = ifelse(month <= 6, # condition
    "first semester", # value if TRUE
    "second semester")) # value if FALSE
```

When we need to differentiate between more cases, we can use `case_when(condition1 ~ new value, condition2 ~ new value, ...)`:

```{r}
# Create a new variable under different conditions with case_when()
txhousing |>
  mutate(trimester = case_when(
    month <= 4 ~ "first trimester",
    4 < month & month <= 8 ~ "second trimester",
    month >= 9 ~ "third trimester"))
```

#### **Try it! Create a new variable called `season` that keeps track of the season in our `txhousing` dataset (in some cases a month could be considered in two different seasons, but only assign this month in which the season starts: e.g., Spring starts in March so March should be in the `spring` category). In which season of which year, was there the highest total number of sales?**

```{r}
# Write and submit code here!
txhousing |>
    mutate(season = case_when(3 <= month & month <= 5 ~ 'Spring',
                              6 <= month & month <= 8 ~ 'Summer',
                              9 <= month & month <= 11 ~ 'Fall',
                              month %in% c(12,1,2) ~ 'Winter')) |>
    slice_max(n = 1, sales)

```

**2015 during the summer in Houston has the highest total number of sales**


We can change the way some values were coded with `recode(old_var, new values in order from least to greatest)`:

```{r}
# Recode a numeric variable into a categorical variable
txhousing |>
  mutate(month_es = recode(month,
                           "enero", "febrero","marzo", 
                           "abril","mayo", "junio",
                           "julio","agosto","septiembre",
                           "octubre","noviembre","diciembre"))
```

Another useful function for analyzing time series in particular is the function `lag()` which refers to a previous value:

```{r}
txhousing |> 
  # Why does it only make sense if we focus on one city at a time?
  filter(city == "Austin") |> 
  # Make sure values are in ascending order per year and month
  arrange(year, month) |>
  # Subtract the previous month's sales with lag(sales))
  mutate(change = sales - lag(sales)) 
```

Let's visualize this difference with a ggplot:

```{r}
txhousing |>
  filter(city == "Austin") |>
  arrange(year, month) |>
  mutate(change_sales = sales - lag(sales)) |> 
  # Let's add a ggplot!
  # Use the variable date on the x-axis (combination of month/year)
  ggplot(aes(x = date, y = change_sales, color = month)) + 
  geom_line() + 
  # Color the lines depending on the month with summer months in red
  scale_color_gradient2(low = "blue", mid = "red", midpoint = 6.5)
  scale_color_grad
```

## 3. Suffixes

We can add suffixes to some core functions such as `summarize` or `mutate`:

-   `_all()`: applies to every variable\
-   `if()`: applies to variables when a conditional statement is TRUE.\
-   `_at()`: applies to some selected variables (using `c("var1", "var2")` or `vars(var1:var3)` for example)\

### a. Summarize

Let's `summarize` with these suffixes! Summarizing all variables with `summarize_all(function, extra argument for that function if needed)`:

```{r}
# Compute the mean for all variables
txhousing |> 
  summarize_all(mean, na.rm = T)
```

However, it doesn't make sense to calculate means for categorical variables and find distinct values for numeric variables. Let's use `summarize_if(condition, function, extra argument for that function if needed)` instead!

```{r}
# Summarize variables that meet certain criteria
txhousing |>
  summarize_if(is.numeric, mean, na.rm = T) # Only compute the mean for numeric variables
```

Or we could be interested in just some particular variables:

```{r}
# Only summarize for a few variables: all of the following are equivalent!
txhousing |>
  summarize_at(c("sales","volume","median"), mean, na.rm = T) # using c()

txhousing |> 
  summarize_at(vars(sales:median), mean, na.rm = T) # using vars()

txhousing |> 
  summarize_at(4:6, mean, na.rm = T) # using indexing
```

### b. Mutate

Let's `mutate` with suffixes!

```{r}
# Coerce all variables into character
txhousing |>
  mutate_all(as.character)
```

```{r}
# Make all character variables as factors
txhousing |> 
  mutate_if(is.character, as.factor)
```

```{r}
# Make some numeric variables as factors
txhousing |> 
  mutate_at(c("year","month"), as.factor)
```

#### **Try it! Using functions with suffixes, find the number of missing values for each variable.**

```{r}
# Write and submit code here!
txhousing |>
    mutate_all(is.na) |>
    summarize_all(sum)

txhousing |>
    summary()
```

------------------------------------------------------------------------

## **Your turn!**

Let's practice our skills to join some Austin weather data in 2014 to the `txhousing` dataset!

1.  We will first upload a dataset containing information about the weather in Austin, TX in 2014. How many observations and variables are there in the dataset? How do you identify a unique observation in this data (i.e., what are the key variables)? *Note: data was retrieved from <https://www.ncei.noaa.gov/products/land-based-station/integrated-surface-database>*

```{r}
# Upload the weather dataset from my github to your environment
atxweather_2014 <- read_csv("https://raw.githubusercontent.com/laylaguyot/datasets/main//ATX_weather_2014.csv")

# Take a look!
head(atxweather_2014)
head(txhousing)

```

**There are 365 observations in the data set with 8 variables. A unique observation is a row with values for all 8 of the variables in the dataset with no duplicates; a unique observation in this case is a particular day so the variables 'date', 'year', 'month', and 'day' show us a unique observation in the dataset**

2.  What would we need to do to join the `weather` data with the `txhousing` data? Change the `weather` and `txhousing` datasets accordingly and save them in your environment with new names.

**Join by 'year', 'month', 'day'. We will need to transform the txhousing dataset according since it only has 'year' and 'month' variables**

```{r}
# Write and submit code here!
tidy_txhousing <- txhousing |>
                    filter(year == 2014 & city == 'Austin')

tidy_atxweather_2014 <- atxweather_2014 |>
                            group_by(year, month) |>
                            summarize_at(c('temperature','pressure', 'precipitation'), mean, na.rm = T)
            
```

3.  Join the information from both datasets so that the joined information makes sense. Using the resulting dataset, create a plot to investigate the impact of the weather on the housing market (i.e., pick one `atxweather_2014` variable and one `txhousing` variable). Add colors and labels to your plot!

```{r}
# Write and submit code here!

atx_housing_and_weather <- full_join(tidy_txhousing, tidy_atxweather_2014, by = c('year', 'month'))

ggplot(data = atx_housing_and_weather) +
    geom_point(aes(x = temperature, y = sales), color = 'purple') +
    labs(title = 'Sales as a function of temperature',
         x = 'temperature (in Farenheit)',
         y = 'number of sales')

```

4.  Post your plot on the following slideshow with your code in the speakers note on bottom: <https://docs.google.com/presentation/d/1XrP0QrpX9V5T3rnytk7_MWe29i_GB5almOuKOfcFTT4/edit?usp=sharing>

------------------------------------------------------------------------

## 4. When to `dplyr` or not `dplyr`

Let's compare a graph after using `dplyr`:

```{r}
txhousing |>
  group_by(year,month) |> # create subsets by year, month
  summarize(mean_vol = mean(volume, na.rm=T), # mean volume
            sd_vol = sd(volume, na.rm=T),     # standard deviation  
            n = n(),                          # number of months in a year
            se_vol = sd_vol/sqrt(n)) |>     # standard error
  ggplot(aes(x = month, y = mean_vol, fill = month)) + 
  geom_bar(stat = "identity") +               # represent the mean volume
  geom_errorbar(aes(y = mean_vol,             # include error bars
                    ymin = mean_vol - se_vol, # define min and max of error bars
                    ymax = mean_vol + se_vol)) +
  facet_wrap(~ year) + 
  scale_x_continuous(breaks = seq(1,12,2)) +  # tick mark for every 2 months
  scale_fill_gradient2(low = "blue", mid = "red", midpoint = 6.5)
```

Versus just a `ggplot`:

```{r}
ggplot(txhousing, aes(x = month, y = volume, fill = month)) + 
  geom_bar(stat = "summary", fun = mean) +                # represent the mean volume
  geom_errorbar(stat = "summary", fun.data = "mean_se") + # represent error bars with SE
  facet_wrap(~ year) + 
  scale_x_continuous(breaks = seq(1,12,2)) +
  scale_fill_gradient2(low = "blue", mid = "red", midpoint = 6.5)
```

Same graph but less code with `ggplot`!


