---
title: "Worksheet 12: String Processing"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE, 
                      warning = TRUE,
                      message = FALSE,
                      fig.align = "center", 
                      R.options = list(max.print=50))

# Edit the file starting below

```

In this worksheet, we will focus on manipulating strings.

## 1. Set up and introduction to the dataset

Let's load `tidyverse` which contains the `stringr` package:

```{r, message=FALSE}
# Load packages 
library(tidyverse)
```

We will refer to a built-in character vector called `fruit` for practicing string manipulations:

```{r}
# Example of a built-in character vector
fruit
```

## 2. Strings

While we have come across some strings before, let's manipulate strings with functions from `stringr`: they have intuitive names, and they all start with `str_`.  

### a. Create strings

We can create strings with either single quotes `'` or double quotes `"`:

```{r}
"Here is an example string"

'Here is another example string'

'Yet another one: Why is this "example" different'
```

To escape the meaning of a special character such as quotes, we can use a backslash `\`:

```{r}
# Single quote
'\''
```

We can store several strings into a vector with `c()`:

```{r}
c("a", "b", "c")
```

Note: `fruit` is a vector of strings.


### b. Manage lengths

Let's use `str_length()` to find the length of a string:

```{r}
# String length
str_length("abc")

str_length("a b c") # a space counts as a character
```

We can apply this function to many strings contained in a vector!

```{r}
# String length of the built-in vector
str_length(fruit)
```

### c. Combine strings

We can use `str_c()` to combine two or more strings:

```{r}
# Combine strings
str_c("Thanksgiving", "Break")

# By default, no space but we can add the argument sep = 
str_c("Thanksgiving", "Break", sep = " ")
```

```{r}
# Add a prefix by using a vector in the second argument
str_c("prefix-", c("a","b","c"))

# Add a suffix by using a vector in the first argument
str_c(c("a","b","c"), "-suffix")

# adding a prefix (first argument) and a suffix (third argument) to the string (second arguement)
str_c("prefix-", c("a","b","c"), "-suffix")
```

What if we want to combine all the values of one vector together?

```{r}
# Use the argument collapse =
str_c(c("a","b","c"), collapse = "")

# Or separate by a comma and a space
str_c(c("a","b","c"), collapse = ", ")
```

#### **Try it! Find the number of letters used in `fruit`, in total.**

```{r}
# Write and submit code here!
str_length(str_c(fruit, collapse='')) # method 1

sum(str_length(fruit)) # method 2

```

**There were a total of 647 characters used to create the fruit string vector**


### d. Mutate strings

We can change the strings from lower to uppercase and vice-versa (also use sentence case):

```{r}
# To lower case
str_to_lower("Thanksgiving Break")

# To upper case
str_to_upper("Thanksgiving Break")

# To sentence case
str_to_sentence("Thanksgiving Break")
```

### e. Subset strings

We can focus on a subset of a string with `str_sub()` (only works with indexing positions though):

```{r}
# Select a position in the string
fruit |> str_sub(start = 1, end = 3)

# Or count backwards with -
fruit |> str_sub(start = -3, end = -1)
```

## 3. Patterns

Let's start recognizing patterns in strings!

### a. Find matches for some patterns

We can find patterns in our data with `str_view()`:

```{r}
# View all possible matches to a pattern within one word 
fruit |> str_view("an")
```

### b. Detect matches for some patterns

What does `str_detect()` do differently?

```{r}
# Detect the matches
fruit |> str_detect("an")
```

#### **Try it! Count how many fruits contain the pattern `berry`.**

```{r}
# Write and submit code here!
sum(str_detect(fruit, 'berry'))

fruit |> str_detect('berry') |> sum()

```

**14 instances of strings with 'berry' in the `fruit` vector**


### c. Subset matches for some patterns

And what about `str_subset()`?

```{r}
# Subset the matches
fruit |> str_subset("an")
```

#### **Try it! Create a vector called `berries` that contains all fruits with the pattern `berry`.**

```{r}
# Write and submit code here!
berries <- str_subset(fruit,'berry')
berries
```

**This creates a vector with only berries**


### d. Split strings

This can be a little tricky. Let's try to split each fruit that has two words into two separate words with `str_split()`:

```{r}
# Split when there is a space
fruit |> str_split(" ")
```

Technically, R created what we call "lists". We can `unlist()` to go back to the format of a vector:

```{r}
# unlist() flattens the vector into a 1D array whereas when we use split() it creates a 2D array

# Split when there is a space
fruit |> str_split(" ") |> unlist()
```

### e. Replace or remove patterns

Let's try `str_replace()` and `str_replace_all()`:

```{r}
# Look for all patterns...
str_view("banana", "an")

# ... and replace one of these patterns
str_replace("banana", "an", "oon")

# ... or replace all of these patterns
str_replace_all("banana", "an", "oon")
```

What about `str_remove()` and `str_remove_all()`?

```{r}
# Remove one of these patterns
str_remove("banana", "an")

# Or remove all of these patterns
str_remove_all("banana", "an")
```

## 4. Regular expressions (Regex)

Regular expressions are used to describe patterns in strings. They're a little weird at first but they can be very useful!

### a. Wildcards

Use `.` to match any character (except a new line):

```{r}
# View all matches with any letter, a, and any letter
str_view(fruit,".a.")
```

### b. Anchors

Let's find a match at the beginning of a string with `^` or at the end of a string with `$` :

```{r}
# Match at the beginning with "a"
fruit |> str_view("^a")
```

```{r}
# Match at the end with "a"
fruit |> str_view("a$")
```

#### **Try it! What is the proportion of fruit names that start with the first letter of your name?**

```{r}
# Write and submit code here!
length(str_view(fruit, '^a')) / length(fruit) * 100
```

**3.75% of fruit names start with the first letter of my name (Austine)**


### c. Literal characters

What if our string include special characters? We need to use an "escape" with backslash. For example, `\\` means literal `\` in strings.

```{r}
# Test weird strings
weirds <- c("a.A", "a\\c", "a/c", "a c") 

# Match literal period
str_view(weirds, "\\.")
```

```{r}
# How is that different from this command?
str_view(weirds, ".\\..") 

# this commands has '.' before and after '\\.' so it matches the character before and after the literal '.' as well
```

```{r}
# Match literal backslash
str_view(weirds, "\\\\")
```

### d. Special patterns

Here are a few useful patterns:

-   `[0-9]` matches any digit

-   `[ ]` matches any single space

-   `[^ ]` matches anything except a space

-   `[abc]` matches a, b, or c

-   `[^abc]` matches anything except a, b, or c

-   `[a-zA-Z]` matches any letter, lower case or upper case

-   `[a-zA-Z0-9]` matches any alphanumeric character

```{r}
# What do these next few lines do? 
str_view(weirds,"[ ]")

# this finds if the string has a space character in it
```

```{r}
str_view(weirds,"[abc]")

# this finds if the string has any of the characters within the []
```

```{r}
str_view(weirds, "[^a-zA-Z]")

# this finds the first match of a non-alphabetical character in the string
```

#### **Try it! Show all vowels in `fruit`.**

```{r}
# Write and submit code here!
str_view(fruit, "[aeiou]")
```


### e. Repetitions

To control how many times a pattern can appear, we can use the following structure:

-   `?` is for 0 or 1 match
-   `+` is for 1 or more matches
-   `*` is for 0 or more matches
-   `{n}` is for exactly n matches
-   `{n,}` is for n or more matches
-   `{,m}` is for at most m matches
-   `{n,m}` is for between n and m matches

```{r}
# Test repeats
repeats <- c("","z","zz","zzz")

# What do these lines do? 
str_view(repeats, "z?")

# this indicates whether there are zero or one occurrences of 'z' in the string
```

```{r}
# Compared to
str_view(repeats, "z+")

# this indicates whether there are one or more of 'z' in that string
```

```{r}
# Compared to
str_view(repeats, "z*")

# this indicates whether there are zero or more matches of 'z' in the string
```

```{r}
# Compared to
str_view(repeats, "z{2}")

# this indicates if there are exactly 2 matches of that character in the string and disregards if there are more
```


------------------------------------------------------------------------

## **Your turn!**

Let's practice some string manipulation with another built-in character vector:

```{r}
# Consider this other built-in character vector
head(sentences)
```

Notice some of the words in `sentences` are fruits. Let's find which sentences mention fruits!

1.  We will consider all fruit names from `fruit` and look for matches in the `sentences`. Combine all the elements from `fruit` into a single string called `testfruit` that contains all the possible names separated by `|`. What does `|` mean in R anyway?

```{r}
# Write and submit code here!
testfruit <- str_c(fruit, collapse = '|')
```

**The '|' means logical OR in R**


2.  Use `str_subset()` to find a subset of `sentences` containing a fruit name. Does there appear to be any issue? Look at the third sentence for example: is there any fruit in that sentence? How could we fix that?

```{r}
# Write and submit code here!
str_subset(sentences, testfruit)
```

**The issues appears to be that it is finding the substring of certain words that are fruit. An example is pearl which contains the substring 'pear' and fight which contains the substring 'fig', both of which are in the `fruit` vector**


3.  Let's try a different approach: split the sentences as words then look for exact matches. First, get rid of punctuation in the sentences with `str_remove_all()`. *Hint: punctuation is anything that is not a letter, a number, or a space.* Then convert all characters as lower case and split `sentences` into individual words with `str_split(" ") |> unlist`. Save the resulting vector as `sentences_words`. What would happen if you don't do `unlist`?

```{r}
# Write and submit code here!
sentences_words <- sentences |>
                        str_to_lower() |>
                        str_remove_all('[^a-zA-Z0-9[ ]]') |>
                        str_split(' ') |>
                        unlist()
```

**If we did not use the unlist() function, then it would have became a 2D where each index of the `sentences` contains a vector of words of the sentence originally at that index**


4.  Use the function `which(fruit %in% sentence_words)` to find which fruits appear in `sentence_words`. Is that correct? How could we fix that?

```{r}
# Write and submit code here!
which(fruit %in% sentences_words)
fruit[1]
fruit[31]
fruit[34]
fruit[54]
fruit[58]
fruit[63]

# second method
fruit_words <- fruit |>
                    str_split(' ')|>
                    unlist()
fruit_words
which(fruit_words %in% sentences_words)
fruit_words[1]
fruit_words[5]
fruit_words[6]
fruit_words[11]
fruit_words[21]
fruit_words[35]
fruit_words[39]
fruit_words[48]
fruit_words[60]
fruit_words[64]
fruit_words[69]
fruit_words[72]
fruit_words[85]
fruit_words[90]
```

**This will not work for fruit with 2 words since it will be comparing it to single words of `sentences_words` since they are all split up by the spaces so we would need to split the fruits (I have included a second method to make this better)**