    ---
title: "Worksheet 3: Visualization Principles"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE, 
                      warning = TRUE,
                      message = FALSE,
                      fig.align = "center", 
                      R.options = list(max.print=50))

# Edit the file starting below

```

This semester we will learn how to make different types of visualizations to explore our data. In this worksheet, we will get familiar with principles of `ggplot`. 

## 1. Dataset

We will consider the built-in dataset `quakes` again. It contains information about earthquakes occurring near Fiji islands since 1964. To refresh our memory, let's take a look:

```{r}
# Take a look at the data frame
head(quakes)
```

We will keep exploring this dataset, using `ggplot` so we need to upload the library `tidyverse` containing all related functions:

```{r}
# Upload the library
library(tidyverse)

# Note: If you are using RStudio ON YOUR OWN COMPUTER AND HAVE NEVER USED ggplot2, you will need to install it first using install.packages("tidyverse")
```

## 2. Create a `ggplot`

Let's first represent where the earthquakes occurred by representing longitude vs latitude.

### a. Define a plot

The `ggplot()` function helps us build a graph. Within this function, we specify the dataframe to explore:

```{r}
# Create a ggplot
ggplot(data = quakes)
```

Well, it's empty because we haven't specified how to represent the variables in the `quakes` object...

### b. Add a geometric object

The `geom_...()` functions helps us define how to represent the variables with geometric objects (points, lines, bars, etc.). For example, `geom_point()` can represent our data with points:

```{r, eval = FALSE}
# Create a ggplot with a geometric object
ggplot(data = quakes) +
  geom_point()
```

Well, it does not work because we haven't specified which variables to represent...

### c. Define the mapping

Using `aes()`, aesthetics, we can define the mapping of the variables to each aesthetic of the plot. We can either define the aesthetics for each geometric object or for the whole plot:

```{r}
# Create a ggplot with mapping variables to a geometric object
ggplot(data = quakes) +
  geom_point(aes(x = long, y = lat))

# which here is equivalent to:
ggplot(data = quakes, aes(x = long, y = lat)) +
  geom_point()

# Note the importance of placing "+" correctly
```

You can add many layers to the `ggplot` and it is a good practice to put the new layers on a new line (be sure to end a line with +).

```{r}
# Add layers to the plot
ggplot(data = quakes, aes(x = long, y = lat)) + 
  geom_point() + 
  geom_line() 

# Note that we would have to define the mapping for each geom if not specified as a whole
ggplot(data = quakes) + 
  geom_point(aes(x = long, y = lat)) + 
  geom_line(aes(x = long, y = lat)) # what happens if we don't define aesthetics?
```

Note that graphs can be saved as an R object and manipulated further:

```{r}
# Save a version of our graph
myplot <- ggplot(data = quakes, aes(x = long, y = lat)) + 
  geom_point()

# Display the graph
myplot
```

Then we can add layers if we would like: 

```{r}
# Add some layers to our saved graph
myplot + geom_line()
```

We can change the appearance of some characteristics of the geometric object directly when we define this object. Let's change the color of the points to make them all blue:

```{r}
# Create a ggplot with mapping variables to a geometric object
ggplot(data = quakes, aes(x = long, y = lat)) +
  geom_point(color = "blue") # blue points
```

Other options such as `size =` and `alpha =` can be applied within a geometric object. Both of these options take numeric values.

#### **Try it! Input numeric values for `size =` and `alpha =` in `geom_point()`. What do these options control?**

```{r}
# Write and submit code here!
myplot + geom_point(size = 5, alpha = 0.1)
```

**Write sentences here!**


### d. Try more mapping

We can add color and other types of aesthetics depending on the values of some variables. For example, let's control the size of the points by the magnitude of the earthquake:

```{r}
# Map the color to the mag variable 
ggplot(data = quakes, aes(x = long, y = lat, color = mag)) +
  geom_point()

# Or equivalent
ggplot(data = quakes, aes(x = long, y = lat)) +
  geom_point(aes(color = mag))
```

We will talk later about color palettes.

#### **Try it! Instead of coloring the points by magnitude, color the points by depth. Anything you notice in this graph?**

```{r}
# Write and submit code here!
ggplot(data = quakes, aes(x = long, y = lat)) + geom_point(aes(color = depth))
```

**Write sentences here!**
The quakes with the lowest depth is concentrated in one area

In the last worksheet, we created a new variable to separate each trench:

```{r}
# Create a new variable
mutate(quakes, trench = ifelse(quakes$long < 175,
                         "New Hebrides", "Tonga and Kermadec"))
```

Let's use a pipe `|>` to create the `trench` variable in `quakes` then use another pipe `|>` to create a plot differentiating the color of the points per trench:

```{r}
# Use a pipe to apply a function to the object quakes
quakes |> 
  # Create a new variable called trench
  mutate(trench = ifelse(long < 175, "New Hebrides", "Tonga and Kermadec")) |> 
  # Create a plot
  ggplot() + # does not need to specify dataframe
  geom_point(aes(x = long, y = lat, color = trench))

# How does the legend of color differ from before? Try the shape aesthetics!
```

### e. Add labels and controlling scales

Plots should be easy to interpret and informative labels are a key element in achieving this goal. The `labs()` function provides customized labels for titles, axes, legends, etc.:

```{r}
# Build a plot...
ggplot(data = quakes, aes(x = long, y = lat, color = depth)) + 
  geom_point() +
  # ...and add labels
  labs(title = "Distribution of the depth across the trenches",
       subtitle = "For eathrquakes that occurred near the Fiji Islands since 1964",
       caption = "Data obtained from Dr. John Woodhouse, Dept. of Geophysics, Harvard University",
       x = "Longitude",
       y = "Latitude",
       color = "Depth (km)")
```

We can also control the scales for a better readibility with functions starting with `scale_`:

```{r}
# Build a plot...
ggplot(data = quakes, aes(x = long, y = lat, color = depth)) + 
  geom_point() +
  # ...and adjust scales
  scale_x_continuous(limits = c(160,190), breaks = seq(160, 190, 2)) + # limits between 160 and 190, show tick marks every 2 degrees
  scale_y_continuous(limits = c(-40,-10), breaks = seq(-40, -10, 5)) # limits between -40 and -10, show tick marks every 5 degrees
```

There are two earthquakes near Lat -18 and -16 degrees and Long 177 and 178 degrees, respectively, that may belong to the trench on the left (New Hebrides).  

#### **Try it! Redefine the `trench` variable to include these two earthquakes in New Hebrides. Then create a `ggplot` representing the location of the earthquakes, colored by depth and shaped by trench. While you're at it, add labels and adjust the scales.**

```{r}
# Write and submit code here!
quakes |> 
    mutate(trench = ifelse(long < 178 & lat > -25, 'New Herbrides', 'Tonga and Kermadec')) |> 
    ggplot(aes(x = long, y = lat)) + 
    geom_point(aes(color = depth, shape = trench)) + 
    scale_x_continuous(limits = c(160, 190), breaks = seq(160, 190, 2)) + 
    scale_y_continuous(limits = c(-40, -10), breaks = seq(-40, -10, 5)) + 
    labs(title = 'Depth of the earthquakes from New Herbrides and Tonga/Kermadec', 
         subtitle = 'For earthquakes that occurs near the Fiji islands since 1964',
         x = 'Longitude',
         y = 'Latitude')
```



### f. Facet

Faceting reproduce a graph for each level of another variable (or combination of variables) and are created using functions that start with `facet_`. Let's take a look at the map for three categories of depths:

```{r}
quakes |> 
  # Create a variable called depth_cat
  mutate(depth_cat = ifelse(depth >= 300, 
                            # Value if TRUE
                            "Deep", 
                            # Value if FALSE (so less than 300 km)
                            ifelse(depth < 70, "Shallow", "Intermediate"))) |> 
  # Create a plot
  ggplot() + 
  geom_point(aes(x = long, y = lat, color = depth_cat)) +
  # Facet by trench
  facet_wrap(~depth_cat)
```

#### **Try it! Investigate the relationship between the magnitude and the number of stations that reported the earthquake and compare this relationship across the two trenches. You can use `geom_smooth()` to identify a trend for each trench. Anything you noticed in this graph?**

```{r}
# Write and submit code here!
quakes |> 
    mutate(trench = ifelse(long < 178 & lat > -25, 'New Herbrides', 'Tonga and Kermadec')) |>
    ggplot(aes(x = mag, y = stations)) + 
    geom_point(aes(color = trench, alpha = 0.1)) +
    facet_wrap(~trench)

quakes |> 
    mutate(trench = ifelse(long < 178 & lat > -25, 'New Herbrides', 'Tonga and Kermadec')) |>
    ggplot(aes(x = mag, y = stations)) + 
    geom_smooth(aes(color = trench, alpha = 0.1)) +
    facet_wrap(~trench)
    
```

**Write sentences here!**
The Tonga and Kermadec had a higher number of stations that reported earth quakes, especially for lower magnitude earth quakes

## 3. Choose a color palette and theme

We can customize many aspects of our graphs by hand (colors, scales, background color, grid, ...) or we can use some themes or palettes other than the defaults.

### a. Color palettes

To define our own palette of colors, we can use `scale_color_gradient()` with setting the color for *low* values and another color for *high* values:

```{r}
# Compare values of magnitude (low: not too dangerous, high: more dangerous)
  ggplot(quakes, aes(x = long, y = lat, color = mag)) + 
  geom_point() +
  # Change the color palette
  scale_color_gradient(low = "yellow", high = "darkred")
```

#### **Try it! Display the depth of earthquakes depending on their location and choosing an appropriate color palette.**

```{r}
# Write and submit code here!
ggplot(quakes, aes(x = long, y = lat)) +
    geom_point(aes(color = depth)) +
    scale_color_gradient(low = 'orange', high = 'blue')
```

**Write sentences here!**


Here is a list of some of the discrete palettes that exist:

```{r}
RColorBrewer::display.brewer.all(type="seq")
```

We can use `scale_color_brewer()` when we are mapping a variable to the aesthetic `color =`:

```{r}
quakes |> 
  # Create a variable called depth_cat
  mutate(depth_cat = ifelse(depth >= 300, "Deep", 
                            ifelse(depth < 70, "Shallow", "Intermediate"))) |> 
  # Compare categories of depth
  ggplot(aes(x = long, y = lat, color = depth_cat)) + 
  geom_point() +
  # Change the color palette
  scale_color_brewer(palette = "Oranges")
```

Or we can create our own set of colors!

```{r}
# List of color blind friendly colors
color_blind_friendly <- c("#999999", "#E69F00", "#56B4E9",
                          "#009E73", "#F0E442", "#0072B2",
                          "#D55E00", "#CC79A7")

# Same graph as before but using color blind 
quakes |> 
  # Create a variable called depth_cat
  mutate(depth_cat = ifelse(depth >= 300, "Deep", 
                            ifelse(depth < 70, "Shallow", "Intermediate"))) |> 
  # Compare categories of depth
  ggplot(aes(x = long, y = lat, color = depth_cat)) + 
  geom_point() +
  # Choose colors saved in our object `color_blind_friendly`
  scale_color_manual(values = color_blind_friendly)
```

### b. Themes

There are so many adjustments we can make to our `ggplot` (change background color, color of axis, color of font, ...) and sometimes it is easier to use some predefined themes:

```{r}
# Still same plot
  ggplot(quakes, aes(x = long, y = lat, color = depth)) + 
  geom_point() +
  # Use the default theme_minimal()
  theme_minimal()
```

Visit this website to find a list of all themes available: <https://ggplot2.tidyverse.org/reference/ggtheme.html>
