---
title: "Worksheet 9: Data Transformation"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE, 
                      warning = TRUE,
                      message = FALSE,
                      fig.align = "center", 
                      R.options = list(max.print=50))

# Edit the file starting below

```

In this worksheet, we'll discuss some more `dplyr` functions to manipulate our data.

## 1. Set up and introduction to the dataset

Let's load `tidyverse` which contains the `dplyr` package:

```{r}
# Load a package
library(tidyverse)
```

We will use the dataset `txhousing` for this worksheet. It contains information about the housing market in Texas. Run `?txhousing` in your console for more details.

```{r}
# Take a look at the first few rows of the dataset
head(txhousing) 
```

Get information about the dimensions, types of variables, and some examples of values with `glimpse()`:

```{r}
# Dimensions and structure of the dataset
glimpse(txhousing) 
```

There are `r nrow(txhousing)` rows and `r ncol(txhousing)` in the dataset. What does one row represent? What does one column represent?

Looking at the documentation with `?txhousing`, we are told that the information about the housing market in Texas was provided by the [TAMU real estate center](https://www.recenter.tamu.edu/). The variables are defined as follows:

| Variables       | Description                                                                                    |
|---------------------|---------------------------------------------------|
| `city`          | Name of multiple listing service (MLS) area                                                    |
| `year`, `month` | Year, Month for the housing market data                                                        |
| `sales`         | Number of sales                                                                                |
| `volume`        | Total value of sales                                                                           |
| `median`        | Median sale price                                                                              |
| `listings`      | Total active listings                                                                          |
| `inventory`     | Amount of time (in months) it would take to sell all current listings at current pace of sales |
| `date`          | Date for the housing market data (`year` + `month` / 12)                                       |

Let's manipulate this dataset, with the 6 core `dplyr` functions, including some we already know!

## 2. Operations on rows/observations

Let's consider some `dplyr` functions that apply to the rows/observations of our dataset.

### a. Filter

Use `filter()` to choose rows/observations verifying some conditions:

```{r}
# Filter with multiple criteria
txhousing |>
  filter(city == "Austin", sales <= 1000)
```

```{r}
# Filter to check missing values for one variable
txhousing |>
  # Use `is.na(variable)`
  filter(is.na(sales)) |>
  # Count the rows
  nrow()
```

#### **Try it! How many rows have values for both the `sales` and `volume` variables (meaning they're not missing)?**

```{r}
# Write and submit code here!
txhousing |>
    filter(!is.na(sales), !is.na(volume)) |>
    nrow()
```

**8034 rows have values for sales and volume**

### b. Arrange

Use `arrange()` to sort rows/observations for some variables Default is ascending (from least to greatest or alphabetically for categories) but to sort in the other direction use `desc()`.

```{r}
# Sort by median house price, least-to-greatest 
txhousing |>
  arrange(median) |>
  # Only see the first 6 rows (lowest median sale prices)
  head()
```

```{r}
# Sort by median house price, least-to-greatest 
txhousing |>
  arrange(desc(median)) |>
  # Only see the first 6 rows (highest median sale prices)
  head()
```

### c. Minimum/Maximum values

Let's try `top_n()` vs `slice_max()`/`slice_min()`, and `top_frac()`.

```{r}
# Select top rows (max values) for a variable
txhousing |> 
  # Use `top_n(number of rows, variables)`
  top_n(n = 4, sales)
```

How does it differ from `slice_max`?

```{r}
# Select top percent of rows (max values) for a variable
txhousing |>
  # Use `slice_max(number of rows, variables)`
  slice_max(n = 4, sales)
```

```{r}
# Select bottom rows (min values) for a variable
txhousing |> 
  # Use `top_n(-number of rows, variables)`
  top_n(n = -4, sales)
```

Why did we get more than 4 rows?

```{r}
# Select top percent of rows (max values) for a variable
txhousing |>
  # Use slice_min(number of rows, variables)
  slice_min(n = 4, sales)
```

```{r}
# Select top percent of rows (max values) for a variable
txhousing |>
  # Use `top_frac(proportion of rows, variables)`
  top_frac(n = 0.001, sales)

# Note: for the minimum values add `-` in front of the proportion
```

#### **Try it! When were the lowest 5 median sale prices for Austin? Display them from lowest to highest.**

```{r}
# Write and submit code here!   
txhousing |>
    slice_min(n = 5, median)
```

**Write sentences here!**

## 3. Operations on columns

Let's consider some `dplyr` functions that apply to the columns/variables of our dataset.

### a. Select

Use `select()` to keep or rename a subset of columns/variables.

```{r}
# Select to keep only some variables
txhousing |>
  # Only see 3 variables
  select(city, year, inventory) |>
  # Only see the first 6 rows
  head()
```

```{r}
# Select to keep columns using indexes
txhousing |>
  select(1,2,8) |>
  head()
```

```{r}
# Drop variables using "-"
txhousing |>
  # See all but 3 variables
  select(-city, -year, -inventory) |>
  head()
```

```{r}
# Select and rename...
txhousing |>
  # Use `select()` to rename some variables new_name = old_name
  select(Location = city, 
         Calendar_Year = year, 
         Total_Listings = listings) |>
  head()
```

```{r}
# or just use rename()
txhousing |> 
  rename(Location = city, 
         Calendar_Year = year, 
         Total_Listings = listings) |>
  head()
```

### b. Mutate

Use `mutate()` to create new columns/variables:

```{r}
# Find the mean sale price per row
txhousing |> 
  mutate(mean_price = volume/sales)
```

#### **Try it! What's the difference between the average price as calculated above and the median sale price? Are these two measures the same? Why/Why not?**

```{r}
# Write and submit code here!
txhousing |>
    mutate(mean_price = volume / sales, diff = median - mean_price)
```

**No these measures are not the same because the prices are skewed higher so the mean will be higher than the median**

## 4. Create summaries

Let's consider some `dplyr` functions that can create some summaries for our dataset.

### a. Summarize

Use `summarize()` (or `summarise()` in British!) to calculate summary statistics on columns/variables. Some useful summary functions: `mean()`, `sd()`, `median()`, `IQR()`, `min()`, `max()`, `n()`, `n_distinct()`, ...

```{r}
# Find the mean inventory
txhousing |>
  summarize(mean_inventory = mean(inventory, na.rm = T)) # ignore NA values
```

```{r}
# Add more summaries:
txhousing |>
            # the mean inventory
  summarize(mean_inventory = mean(inventory, na.rm = T), 
            # the median inventory
            median_inventory = median(inventory, na.rm = T),
            # the number of rows  
            n_rows = n(), # number of rows
            # the number of distinct cities in the dataset
            n_cities = n_distinct(city))
```

#### **Try it! Find the total number of `sales` for Austin in 2009 as well as the mean number of sales per month that year.**

```{r}
# Write and submit code here!
txhousing |>
    filter(city == 'Austin', year == '2009') |>
    summarize(total_num_sales = sum(sales, na.rm = TRUE),
              mean_num_sales = mean(sales, na.rm = TRUE))
```

**Write sentences here!**

Now what if we wanted to generate a similar report for all years in `txhousing`? Let's use a function that allows us to create summaries per subgroup.

### b. Group by

This is one very important function! It enables us to create subgroups and apply a function to all these subgroups For example, find summaries per city and per year:

```{r}
# Find summaries by subgroups 
txhousing |>
  # Each year is a subgroup
  group_by(year) |> 
  # Create summaries for each subgroup
  summarize(total_sales = sum(sales, na.rm = TRUE), # total number of sales
            nb_rows = n()) # count how many rows in each subset 
```

Why are there less rows in 2015? How could it influence the total number of sales?

```{r}
# Stop grouping with ungroup()
txhousing |>
  # Each year is a subgroup
  group_by(year) |>
  # Create a summary for each subgroup
  summarize(nb_rows = n()) |>
  # Ungroup to consider all years again
  ungroup() |> 
  # Summarize over all the years: how many of these counts are distinct?
  summarize(n_distinct(nb_rows)) 
```

Let's try to be a little more specific and find the total number of sales per year and per month:

```{r}
# Find summaries by subgroups 
txhousing |>
  # Each year/month is a subgroup
  group_by(year, month) |> 
  # Create summaries for each subgroup
  summarize(total_sales = sum(sales, na.rm = TRUE), # total number of sales
            nb_rows = n()) # count how many rows: what does this number correspond to?
```

After manipulating our dataset, we can build add a `ggplot`:

```{r}
# Let's add a ggplot after using some of the dplyr functions above!
txhousing |>
  group_by(year, month) |> 
  summarize(total_sales = sum(sales, na.rm = TRUE)) |>
  # Add a ggplot
  ggplot(aes(x = month, y = total_sales, color = as.factor(year))) + 
  geom_line() +
  # Change the color palette
  scale_color_ordinal() +
  # Add some labels
  labs(y = "Total monthly sales",
       title = "Monthly sales over the years")
```

Looks like sales get hot in the summer!

------------------------------------------------------------------------

## **Your turn!**

Let's practice some `dplyr` functions on the dataset `txhousing` (think logically).

1.  Is anyone in your group from Texas? Are the cities where your group members are from in the `txhousing` dataset?

```{r}
# Write and submit code here!
txhousing |>
    filter(city == 'Austin' | city == 'Dallas' | city == 'Houston') |>
    group_by(city) |>
    summarize(n())
```

**Austin, Dallas and Houston are in the data set**

2.  What month were you born?

**JANUARY**

3.  Filter rows that contain the cities found in 1. and the months found in 2. *If no one is from Texas, filter 2-3 cities of your choice.* Find the mean number of sales for each month and for each city over the years. Create a visualization to investigate if there is a difference in the number of sales on average between the different cities/between these months. Add colors and labels to your plot!

```{r}
# Write and submit code here!
txhousing |>
    filter(city %in% c('Austin', 'Dallas', 'Houston'), month %in% c(1, 6)) |>
    group_by(city, month) |>
    summarize(mean_num_sales = mean(sales, na.rm = TRUE)) |>
    ggplot(aes(x = month, y = mean_num_sales, fill = city)) +
    geom_bar(stat = 'identity', position = 'dodge') +
    scale_x_continuous(breaks = c(1,6))
    labs(y = 'mean number of sales')
```

**Write sentences here!**

4.  Post your plot on the following slideshow with your code in the speakers note on bottom: <https://docs.google.com/presentation/d/1oQn2HURQkez2UOJnI5DG98rhIcAC1o-XE6r-gcig7m4/edit?usp=sharing>
